\chapter{How the Portuguese made a Beautiful Auto-da-fé, to prevent any further Earthquakes: and how Candide was publicly whipped}
\lettrine[lraise=0.1,nindent=0em,slope=-.5em]{A}{fter} the earthquake had destroyed three-fourths of Lisbon, the sages of that country could think of no means more effectual to prevent utter ruin than to give the people a beautiful \textit{auto-da-fe}\footnote{This \textit{auto-da-fe} actually took place, some months after the earthquake, on June 20, 1756.}; for it had been decided by the University of Coimbra, that the burning of a few people alive by a slow fire, and with great ceremony, is an infallible secret to hinder the earth from quaking.

In consequence hereof, they had seized on a Biscayner, convicted of having married his godmother, and on two Portuguese, for rejecting the bacon which larded a chicken they were eating\footnote{The rejection of bacon convicting them, of course, of being Jews, and therefore fitting victims for an \textit{auto-da-fe}.}; after dinner, they came and secured Dr. Pangloss, and his disciple Candide, the one for speaking his mind, the other for having listened with an air of approbation. They were conducted to separate apartments, extremely cold, as they were never incommoded by the sun. Eight days after they were dressed in \textit{san-benitos}\footnote{The \textit{San-benito} was a kind of loose over-garment painted with flames, figures of devils, the victim's own portrait, etc., worn by persons condemned to death by the Inquisition when going to the stake on the occasion of an \textit{auto-da-fe}. Those who expressed repentance for their errors wore a garment of the same kind covered with flames directed downwards, while that worn by Jews, sorcerers, and renegades bore a St. Andrew's cross before and behind.} and their heads ornamented with paper mitres. The mitre and \textit{san-benito} belonging to Candide were painted with reversed flames and with devils that had neither tails nor claws; but Pangloss's devils had claws and tails and the flames were upright. They marched in procession thus habited and heard a very pathetic sermon, followed by fine church music. Candide was whipped in cadence while they were singing; the Biscayner, and the two men who had refused to eat bacon, were burnt; and Pangloss was hanged, though that was not the custom. The same day the earth sustained a most violent concussion.

Candide, terrified, amazed, desperate, all bloody, all palpitating, said to himself:

``If this is the best of possible worlds, what then are the others? Well, if I had been only whipped I could put up with it, for I experienced that among the Bulgarians; but oh, my dear Pangloss! thou greatest of philosophers, that I should have seen you hanged, without knowing for what! Oh, my dear Anabaptist, thou best of men, that thou should'st have been drowned in the very harbour! Oh, Miss Cunegonde, thou pearl of girls! that thou should'st have had thy belly ripped open!''

Thus he was musing, scarce able to stand, preached at, whipped, absolved, and blessed, when an old woman accosted him saying:

``My son, take courage and follow me.''

